BIOST 578 HW 1  
========================================================


The goal of this homework is to find all HCV gene expression data using the Illumina platform submitted by an investigator at Yale.

First, install from Bioconductor the `GEOmetadb` package, which contains all metadata associated with the GEO database.  

```{r, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("GEOmetadb")
```

Then load the package. 

```{r, message=FALSE}
library("GEOmetadb")
```

We also need to initiate a local SQLite database to store the metadata associated with the GEO database.

```{r eval=FALSE}
getSQLiteFile()  # only need to do this once 
```

Then fill in the SQLite database with the GEO database metadata. 

```{r}
geo_con <- dbConnect(SQLite(),'GEOmetadb.sqlite')
```

# Using the `dbGetQuery` function

Now we can start quering the database to get information on the dataset that we want. We will use the `dbGetQuery` function to do the query. Specifically we are looking for entrees that have "HCV" in the title cell, "Illumina" in the manufacturer cell and "Yale" in the contact cell. The output is the title, GSE accession number, GPL accession number and manufacturer and description of platform for entrees that match. 

```{r}
dbGetQuery(geo_con, "SELECT gse.title, gse.gse, gpl.gpl, gpl.manufacturer, gpl.technology FROM (gse JOIN gse_gpl ON gse.gse=gse_gpl.gse) j JOIN gpl ON j.gpl=gpl.gpl WHERE gse.summary LIKE '%HCV%' AND gpl.manufacturer LIKE '%Illumina%' AND gse.contact LIKE '%Yale%';")
```

# Using `data.table`s

Alternatively, we can convert the relevant database tables into data tables and query the data tables instead for the data that we want.

First, load the `data.table` package. 

```{r, message=FALSE}
library(data.table)
```

Convert the gse, gpl and gse_gpl database tables into data tables.

```{r}
gse <- data.table(dbReadTable(con=geo_con, name="gse", row.names=FALSE, header=TRUE, sep="\t"))
gpl <- data.table(dbReadTable(con=geo_con, name="gpl", row.names=FALSE, header=TRUE, sep="\t"))
gse_gpl <- data.table(dbReadTable(con=geo_con, name="gse_gpl", row.names=FALSE, header=TRUE, sep="\t"))
```

We need to merge the gse, gpl and gse_gpl data tables into one table before we can extract the information we want. Because some of the column names in the gse and gpl tables are the same, to avoid confusion after merging the tables, we will prefix the column names in each table by the name of the table (e.g. change the ID column name in the gse table into gse.ID).

```{r}
setnames(gse, apply(as.matrix(colnames(gse)), MARGIN=1, FUN=function(x) {paste0("gse.", x)}))
setnames(gpl, apply(as.matrix(colnames(gpl)), MARGIN=1, FUN=function(x) {paste0("gpl.", x)}))
```

The merging is a two-step process. First we merge the gse and gse_gpl tables using the gse column as a key. Then we merge the resulting table with the gpl table using the gpl column as a key.

```{r}
## Set keys to join gse and gse_gpl tables
setkey(gse, "gse.gse")
setkey(gse_gpl, "gse")

## Inner join gse and gse_gpl tables
j <- gse[gse_gpl, nomatch=0]

## Set keys to join j and gpl tables
setkey(j, "gpl")
setkey(gpl, "gpl.gpl")

## Inner join j and gpl tables
k <- gpl[j, nomatch=0]
```

Finally proceed with the query. This query is equivalent to the query we did using the `dbGetQuery` function.

```{r}
k[gse.summary %like% "HCV" & gpl.manufacturer %like% "Illumina" & gse.contact %like% "Yale", list(gse.title, gse.gse, gpl.gpl, gpl.technology)]
```

Note that the query result is the same as `dbGetQuery`.